---
title: "DATA622: Homework 3"
author: "by Thomas Hill"
output:

  html_document:
    highlight: pygments
    number_sections: no
    theme: cerulean
    toc: yes
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, error=FALSE, warning=FALSE, message=FALSE, fig.align = "center")
```

Perform an analysis of the dataset used in Homework #2 using the SVM algorithm. Compare the results with the results from previous homework.

```{r, libraries}

library(e1071)
library(dplyr)
library(caret)
library(caTools)

```


```{r, csv}

aaa_df <- read.csv('~/GitHub/DATA622/Absenteeism_at_work.csv', sep = ';', colClasses=c(rep("factor",5), rep('numeric', 6), rep('factor', 2), 'numeric', rep('factor', 2), rep('numeric', 5)))
```


For this week's homework, I'm using UCI ML's 'Absenteeism' data set. This data set has information about callouts in a Brazilian firm over a 3 year period.

```{r, preprocessing}
aaa_df[738,]$Month.of.absence = factor(7) #impute to most common months
aaa_df[739,]$Month.of.absence = factor(3)
aaa_df[740,]$Month.of.absence = factor(4)
aaa_df <- aaa_df %>%
  group_by(Seasons) %>%
  mutate(Month.of.absence = replace(Month.of.absence, Month.of.absence == 0, 7)) #impute to most common month
aaa_df[52,]$ID = factor(28) #correct ID
```






```{r, train-test-split}
set.seed(0424)

sample <- sample.split(1:nrow(aaa_df), SplitRatio = 0.8)
train <- subset(aaa_df, sample == TRUE)
test <- subset(aaa_df, sample == FALSE)
```


```{r, scale-dummies-train}

scale <- sapply(train, function(i) if(is.numeric(i)) scale(i) else i)

dummies <- dummyVars(~., scale[,-20])

d_predict <- predict(dummies, train[,-20])
dummies_train <- as.data.frame(cbind(train$Absenteeism.time.in.hours, d_predict)) %>%
  select(-Month.of.absence0) #this factor level is not used

```


```{r, scale-dummies-test}

scale <- sapply(test, function(i) if(is.numeric(i)) scale(i) else i)

dummies <- dummyVars(~., scale[,-21])

d_predict <- predict(dummies, test[,-21])
dummies_test <- as.data.frame(cbind(test$Absenteeism.time.in.hours, d_predict)) %>%
  select(-Month.of.absence0) #this factor level is not used

```



```{r, svm-model}

s_mod1 <- svm(dummies_train[,-1], dummies_train$V1)


svm1 <- predict(s_mod1, dummies_test[,-1])

summary(svm1)
```

```{r, svm-performance}

summary(s_mod1)

mean(s_mod1$residuals^2)
```

The performance of SVM on these training data is better than the random forest model in Homework 2 but slightly worse than the more complicated decision tree model







